% me=0 student solutions (ps file), me=1 - my solutions (sol file),
% me=2 - assignment (hw file)
\def\me{0} \def\num{11} %homework number

\def\due{5 pm on Tuesday, December 05} %due date

\def\course{CSCI-GA.1170-001, 003 Fundamental Algorithms} %course name, changed only once

% **** INSERT YOUR NAME HERE ****
\def\name{Atmaj Koppikar}

% **** INSERT YOUR NETID HERE ****
\def\netid{ark9238}

% **** INSERT NETIDs OF YOUR COLLABORATORS HERE ****
\def\collabs{NetID1, NetID2}


\iffalse

INSTRUCTIONS: replace # by the homework number.  (if this is not
ps#.tex, use the right file name)

Clip out the ********* INSERT HERE ********* bits below and insert
appropriate LaTeX code.  There is a section below for student macros.
It is not recommended to change any other parts of the code.


\fi
%

\documentclass[11pt]{article}

% ==== Packages ====
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{latexsym}
\usepackage{etoolbox}
\usepackage{totcount}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage[bottom]{footmisc}
\usepackage{enumitem}
\usepackage{hyperref}

\usepackage{array}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{arrows}

\setlength{\footskip}{1in} \setlength{\textheight}{8.5in}

\newcommand{\handout}[5]{
	\renewcommand{\thepage}{#1, Page \arabic{page}}
	\noindent
	\begin{center}
		\framebox{ \vbox{ \hbox to 5.78in { {\bf \course} \hfill #2 }
				\vspace{4mm} \hbox to 5.78in { {\Large \hfill #5 \hfill} }
				\vspace{2mm} \hbox to 5.78in { {\it #3 \hfill #4} }
				\ifnum\me=0
				\vspace{2mm} \hbox to 5.78in { {\it Collaborators: \collabs
						\hfill} }
				\fi
		} }
	\end{center}
	\vspace*{4mm}
}

\newcounter{pppp}
\newcounter{pppc}
\newcommand{\prob}{\arabic{pppp}} %problem number
\newcommand{\increase}{\stepcounter{pppp}\stepcounter{pppc}} %problem number

% Arguments: Title, Number of Points
\newcommand{\newproblem}[2]{
	\ifnum\me=0
	\ifnum\prob>0 \newpage \fi
	\increase
	\restartlist{subtasks}
	\setcounter{page}{1}
	\handout{\name{} (\netid), Homework \num, Problem \arabic{pppp}}
	{\today}{Name: \name{} (\netid)}{Due: \due}
	{Solutions to Problem \prob\ of Homework \num\ (#2)}
	\else
	\increase
	\restartlist{subtasks}
	\section*{Problem \num-\prob~(#1) \hfill {#2}}
	\fi
}

\newlist{subtasks}{enumerate}{1}
\setlist[subtasks]{label={(\alph*)},resume}

\newcounter{numpppp}
\loop
\stepcounter{numpppp}
% workaround bug in totcount (means \newtotcounter{points\arabic{pointsct}})
\begingroup%
\edef\tempcounter@@name{points\arabic{numpppp}}%
\expandafter\newtotcounter\expandafter{\tempcounter@@name}%
\edef\tempcounter@@name{ecpoints\arabic{numpppp}}%
\expandafter\newtotcounter\expandafter{\tempcounter@@name}%
\endgroup
\ifnum \value{numpppp}<500 % max number of tasks supported
\repeat

% formating of output
\newcommand{\disppoints}[1]{%
	\texorpdfstring{(#1~\ifnumequal{#1}{1}{point}{points})}{}
}

% adds and displays
\newcommand{\points}[1]{%
	\texorpdfstring{\addtocounter{points\arabic{pppc}}{#1}\disppoints{#1}}{}%
}
\newcommand{\ecpoints}[1]{%
	\texorpdfstring{\addtocounter{ecpoints\arabic{pppc}}{#1}\ec \disppoints{#1}}{}%
}
\newcommand{\mixedpoints}[2]{%
	\texorpdfstring{%
		\addtocounter{points\arabic{pppc}}{#1}%
		\addtocounter{ecpoints\arabic{pppc}}{#2}%
		(#1 (+#2) points)
	}{}%
}

% total points of current task         
\newcommand{\currentpoints}{% total points of current task   
	\texorpdfstring{\ifnumequal{\totvalue{ecpoints\arabic{pppc}}}{0}%
		{\total{points\arabic{pppc}} points}%
		{\total{points\arabic{pppc}}+\total{ecpoints\arabic{pppc}} points}}{}%
}


\def\squarebox#1{\hbox to #1{\hfill\vbox to #1{\vfill}}}
\def\qed{\hspace*{\fill}
	\vbox{\hrule\hbox{\vrule\squarebox{.667em}\vrule}\hrule}}
\newenvironment{solution}{\begin{trivlist}\item[]{\bf Solution:}}
	{\qed \end{trivlist}}
\newenvironment{solsketch}{\begin{trivlist}\item[]{\bf Solution
			Sketch:}} {\qed \end{trivlist}}
\newenvironment{code}{\begin{tabbing}
		12345\=12345\=12345\=12345\=12345\=12345\=12345\=12345\= \kill }
	{\end{tabbing}}


\newcommand{\hint}[1]{({\bf Hint}: {#1})}
% Put more macros here, as needed.
\newcommand{\room}{\medskip\ni}
\newcommand{\brak}[1]{\langle #1 \rangle}
\newcommand{\bit}[1]{\{0,1\}^{#1}}
\newcommand{\zo}{\{0,1\}}
\newcommand{\C}{{\cal C}}

\newcommand{\nin}{\not\in}
\newcommand{\set}[1]{\{#1\}}
\renewcommand{\ni}{\noindent}
\renewcommand{\gets}{\leftarrow}
\renewcommand{\to}{\rightarrow}
\newcommand{\assign}{:=}

\newcommand{\AND}{\wedge}
\newcommand{\OR}{\vee}
\renewcommand{\And}{\mbox{\bf and }}
\newcommand{\Or}{\mbox{\bf or }}

\newcommand{\For}{\mbox{\bf for }}
\newcommand{\To}{\mbox{\bf to }}
\newcommand{\DownTo}{\mbox{\bf downto }}
\newcommand{\Do}{\mbox{\bf do }}
\newcommand{\If}{\mbox{\bf if }}
\newcommand{\Then}{\mbox{\bf then }}
\newcommand{\Else}{\mbox{\bf else }}
\newcommand{\Elseif}{\mbox{\bf else if }}
\newcommand{\While}{\mbox{\bf while }}
\newcommand{\Repeat}{\mbox{\bf repeat }}
\newcommand{\Until}{\mbox{\bf until }}
\newcommand{\Return}{\mbox{\bf return }}
\newcommand{\Halt}{\mbox{\bf halt }}
\newcommand{\Swap}{\mbox{\bf swap }}
\newcommand{\Ex}[2]{\textrm{exchange } #1 \textrm{ with } #2}
\newcommand{\Nil}{\mbox{\bf nil}}
\newcommand{\In}{\mathsf{inOrder}}
\newcommand{\Post}{\mathsf{postOrder}}
\newcommand{\Pre}{\mathsf{preOrder}}
\newcommand{\Root}{\mathsf{root}}
\newcommand{\Parent}{\mathsf{parent}}
\newcommand{\Left}{\mathsf{left}}
\newcommand{\Right}{\mathsf{right}}
\newcommand{\Middle}{\mathsf{middle}}
\newcommand{\True}{\textbf{true}}
\newcommand{\False}{\textbf{false}}
\newcommand{\Print}{\mbox{\bf print }}
\newcommand{\ec}{({\bf Extra Credit})}
\newcommand{\note}{{\bf Note to Graders: }}
\newcommand{\Note}{\note}
\newcommand{\Rotate}{\textsc{Rotate}}
\newcommand{\LRotate}{\textsc{LeftRotate}}
\newcommand{\RRotate}{\textsc{RightRotate}}
 \newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\Gray}{\textsc{gray}}
\newcommand{\White}{\textsc{white}}
\newcommand{\Black}{\textsc{black}}

\begin{document}





\ifnum\me=0

% Collaborators (on a per task basis):
%
% Task 1: *********** INSERT COLLABORATORS HERE *********** 
% Task 2: *********** INSERT COLLABORATORS HERE *********** 
% etc.
%

\fi

\ifnum\me=1

\handout{PS \num}{\today}{Lecturer: Yevgeniy Dodis}{Due: \due}
{Solution {\em Sketches} to Problem Set \num}

\fi

\ifnum\me=2

\handout{PS \num}{\today}{Lecturer: Yevgeniy Dodis}{Due: \due}{Problem
  Set \num}

\fi




\newproblem{Greedy Topological Sort}{\currentpoints}


Let $G = (V,E)$ be a directed graph and let $n=|V|$ and $m=|E|$ denote
the number of vertices and edges, respectively.
Consider the following greedy algorithm for topological
sort of $G$: 
``Find a vertex $v$ with no outgoing
edges. If no such $v$ exists, output `cyclic'. Else put $v$ as the
last vertex in the topological sort, remove $v$ from $G$ (by also
removing all incoming edges to $v$), and recurse on the remaining
graph $G'$ on $(n-1)$ vertices. For the empty graph, end the algorithm.''.



In a first part, we will prove the above algorithm to be correct.

\begin{subtasks}
	
	\item \points{3}
	Let $G$ to be acyclic with $n > 0$, i.e., non-empty. Show that $G$ has at least
	one vertex $v$ having no outgoing edges. 
	\hint{Do a proof by contradiction.}
	
	\ifnum\me<2
\begin{solution}
		An acyclic directed graph can be topologically sorted. As in in the nodes in the graph can be arranged in a line such that all edges
		move from left to right. eg. $A->B->C->D B->E$ In this case vertices D and E have no outgoing edges.

		If any of them did have outgoing edges that point towards a node to their left then that would cause a cycle. eg. $A->B->C->D B->E E->A$
		If E did not have any outgoing edges then the graph would be acyclic. It is possible for E to have an edge and still be acyclic. eg. $A->B->C->D B->E E->C$.
		 In this case only D has no outgoing edges and still the graph is acyclic. Addingg an edge from D to any other edge would cause a cycle.

		This means that at least one vertex must have no out-going edges for a graph to be acyclic. 
\end{solution}
	\fi
	
	
	\item \points{2}
	Now prove that the algorithm produces output ``cyclic'' if an only if there 
	is no topological sorting. \hint{Use part~(a) as part of an induction proof over the number of vertices $n$.}
	
	\ifnum\me<2
\begin{solution} 
As given in part a, a vertex in an acyclic graph is placed to the right of vertices that point to it in the topological order. 
The rightmost vertex has no outgoing edges. 

Base Case :

A graph with one vertex is acyclic. In topological order the last vertex has no outgoing edges.

Inductive CAse : 

Assuming that a topologically sorted graph has no outgoing edges from the last vertex.

For every new node added to the sort, if it has no outgoing edges to any vertex already in the sort, then the graph is topologically sorted.
 Iterate through every vertex in the graph and if it satisfies the condition then add it to the sort.

AFter processing all the vertices if the sort has vertex with no outgoing edges and it is placed last then the graph is acyclic.

Therefore the algorithm will only output cyclic if and only if there is no vertex with no outgoing edges because then the topological sort would not be possible.
\end{solution}
	\fi
	
	
	\item \points{2}
	Now assume $G$ to be acyclic. Conclude the correctness proof by showing that the algorithm outputs a valid topological sorting.
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}


In the following, we want to efficiently implement to above algorithm.
To this end, assume $G = (V,E)$ to be given in adjacency-list format, where each
adjacency list stores the \emph{outgoing} edges.

\begin{subtasks}
	\item \points{3}
	Show how to implement to above algorithm in time $O(n (m+n))$.
	\hint{Assume that you can set flags $v.color$ on each node $v \in V$ indicating whether the 
		node has already been output, i.e., removed from the graph. Do not try to explicitly construct $G'$ for the recursive invocation.} 
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
	
	
	\item \points{4}
	Now assume that the graph $G$ is given in an adjacency-list format that stores the \emph{incoming} edges instead of the outgoing ones. (Converting between those two formats can be done in time $O(n+m)$.) 
	Show how to implement the algorithm in time $O(n^2)$ in this case. 
	\hint{You may want to preprocess the graph and store some information at each node that then gets updated through the algorithm.} 
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}



\newproblem{Cool Orderings}{\currentpoints}

Recall that in the algorithm for computing the strongly connected
components (SCCs) of a graph $G = (V,E)$, the second DFS on the transpose
$G^T$ (i.e., on the graph with reversed edges) processes the nodes in decreasing order of their finishing time (w.r.t.\ the first DFS).

Define an ordering $v_1,\ldots,v_n$ of the vertices of $G$ to be
\emph{cool} if
\begin{itemize}
	\item the vertices can be split up into $k$ groups corresponding to
	the $k$ SCCs of $G$ by appropriately inserting $k-1$ separators,
	i.e., if there exist $i_1,\ldots,i_{k-1}$ such that the SCCs are
	$\set{v_1,\ldots,v_{i_1}},\set{v_{i_1+1},\ldots,v_{i_2}},\ldots,
	\set{v_{i_{k-1}+1},\ldots,v_n}$, and
	\item the $k$ groups are in topological order (w.r.t.\ the SCC
	graph).
\end{itemize}
For example, assume $G$ has vertices $V = \set{a,b,c,d,e}$ and edges
$E= \set{(a,c),(c,a),(a,d)} \cup \set{(b,e),(e,b),(b,d)}$.  Then, $G$
has SCCs $C_1 = \set{a,c}$, $C_2 = \set{b,e}$, and $C_3 = \set{d}$.
Moreover, the SCC graph of $G$ has edges $\set{(C_1,C_3),(C_2,C_3)}$.
Hence, $a,c,e,b,d$ and $b,e,a,c,d$ are examples of cool orderings,
whereas $b,a,c,d,e$ and $a,b,c,d,e$ are uncool.

\begin{subtasks}
	\item \points{3}
	Prove or disprove: sorting the vertices in decreasing order of their finishing time (of an arbitrary DFS) is always cool.
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}
In the following, we want to prove that if the DFS is run on $G^T$ 	w.r.t.\ some cool ordering, the resulting DFS trees are the SCCs of $G$. To this end, we adapt the proof from the book accordingly.
Recall that $f(v)$ denotes the finishing time of node $v$ and for a component $U \subseteq V$, $f(U) = \max_{v \in U} f(v)$ denotes the last finishing time of any node in the component.

\begin{subtasks}
	\item \points{4} 
	Adapt the following lemma from the book to be about cool orderings instead of finishing times and prove the resulting lemma.
	\begin{lemma}[Lemma 20.14]
		Let $C$ and $C'$ be distinct strongly connected components in directed graph $G = (V,E)$. Suppose that there is an edge $(u,v) \in E$, where $u \in C'$ and $v \in C$. Then $f(C') > f(C)$.
	\end{lemma}
	\hint{Replace the function $f$ with an appropriate function $g$ and adjust the lemma where as needed.}
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}

\noindent
Assuming you have solved subtask (b) correctly, one can adapt Corollary 20.15 and Theorem 20.16 verbatim (replacing $f$ with $g$) and obtain a proof that using a cool ordering indeed works for finding the SCCs. 






\newproblem{Flights of Fancy}{\currentpoints}


You are hired as the technical
consultant for an airlines company
called ``Untied'' to help improve
their connectivity and also their
profit margins. You are given
a weighted, undirected graph $G=(V,E)$ where
each node represents the cities
where Untied serves and an
edge $(u,v)$ denotes that there
is a flight service from $u$
to $v$ and from $v$ to $u$. Let $|V|=n,|E|=m$. 
The graph
need not be connected. In
parts (a)-(e), the weight corresponds
to the cost of operating that route. 

In a preliminary survey, you identify
a potential problem that is causing
reduced profit: Some customers have managed to find
hacker-fares which exploit indirect routings. That is,
they manage to bypass find cheaper fares from $u$ to $v$
(where a direct flight exists) via a sequence of one or more
intermediate stops. Therefore, you are asked to figure out 
whether your route network is a \emph{spanning tree}. 

\begin{subtasks}
	\item \points{5}
	Present an $O(n)$ algorithm where $|V|=n$ to determine if
	the given graph $G$ is a spanning tree
	or not. More formally,
	
	\begin{center}
		Input: A weighted, undirected graph $G=(V,E)$
		
		Output: Print $\True$ iff $G$ is a spanning tree,
		else print a possible cycle, i.e., the vertices in this cycle
		
		Constraint: $O(n)$-time algorithm
		
	\end{center}
	
	Clearly, describe your
	algorithm. Do not forget to
	argue the correctness and prove the running time. 
	
	You can invoke any algorithm discussed in
	the lecture, as a black-box. You can also
	modify these algorithms. Clearly state these
	modifications and justify the impact of
	these changes on the running time and correctness
	of the algorithm. 
	\hint{It is easy to define an $O(m+n)$
		algorithm that works even for directed graphs. However,
		here we want you to get an algorithm that
		runs in time $O(n)$ which is a huge save
		if $m>>n$. This algorithm relies on
		the fact that $G$ is undirected.}
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}

\noindent
Now, you have determined that $G$
is not a spanning tree but $G$ is connected.
It makes sense to produce a spanning
tree whose weight is minimal. You can run
Kruskal's algorithm to produce such a spanning
tree. The pseudocode is provided below:

\begin{code}
	{\sc Kruskal}$(G,w)$\\
	1 \> $A=\emptyset$\\
	2 \> \For each vertex $v\in G.V$\\
	3 \> \> {\sc Make-Set}$(v)$\\
	4 \> sort the edges of $G.E$ into non-decreasing order by weight $w$\\
	5 \> \For each $(u,v)$ from the sorted list\\
	6 \> \> \If {\sc Find-Set}$(u)\neq\textsc{Find-Set}(v)$\\	
	7 \> \> \> $A=A\cup\{(u,v)\}$\\
	8 \> \> \> {\sc Uniion}$(u,v)$\\
	9 \> \Return $A$
\end{code}

\begin{subtasks}
	\item \points{6}
	In this question, you are asked to account for another constraint:
	routes cannot be removed willy-nilly, especially because
	of contracts.  
	
	Specifically, you are given
	this set of routes as $F=(V',E')$ which is \emph{a forest of edges}.
	Now, you are asked to produce the best
	possible spanning tree such that $F$ is always included
	in the resulting tree. More formally,
	
	\begin{center}
		Input: A weighted, connected, undirected graph $G=(V,E)$,
		a forest $F=(V',E')$ that is a sub-graph of $G$, i.e., $V' \subseteq V$ and $E' \subseteq E$.
		
		Output: Return the best possible spanning tree subject
		to the constraint below.
		
		Constraint: Each edge in $F$ is included in the
		spanning tree. 
		
	\end{center}
	
	
	Modify the pseudocode {\sc Kruskal}$(G,w)$ to
	{\sc Kruskal}$(G,F,w)$ that produces the new
	spanning tree. You will do this by inserting
	no more than five lines of code to the pseudocode
	provided earlier. Your algorithm should still run in
	$O(|E|\log |V|)$.
	
	Do not forget to argue the correctness and justify the running time. 
	\hint{Use a theorem from the book for correctness.}
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}




\newproblem{Least Maximal-Weight Edge}{\currentpoints}

Recall from the lecture that the minimum-spanning-tree (MST) problem
is to find, in some undirected graph~$G$, a spanning sub-tree $T$ with
the smallest possible sum $\sum_{e\in T} w(e)$ of edge weights.
Consider the related problem of finding a spanning sub-tree $T'$ with
the least maximum edge weight $\max_{e\in T'} w(e)$.

\begin{subtasks}
	\item \points{4}
	Assume the edge weights are unique.  Argue why
	Kruskal's algorithm outputs a spanning tree $T$ that is also a
	(optimal) solution to the modified problem.  You may use the fact
	that Kruskal is a correct algorithm.
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
	
	
	\item \mixedpoints{2}{2}
	Find a graph that contains a non-MST spanning tree $T$ with least maximum edge weight.  
	That is $T$ is a (optimal) solution to the modified problem but not an MST.  For
	\textbf{extra credit}, the edge weights must be unique.
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
	
	\item \points{4} 
	Consider now graphs with edge weights that are not necessarily unique.  
	Show that any (optimal) solution $T$ to the	MST problem is also an (optimal) 
	solution to the modified problem.
	
	\noindent
	\hint{Do not use a specific algorithm's (such as Kruskal's or
		Prim's) correctness since they output some specific MST out of
		possibly many. Give a general argument instead.}
	
	\ifnum\me<2
\begin{solution}   INSERT YOUR SOLUTION HERE   \end{solution}
	\fi
\end{subtasks}





\end{document}
